---
title: "Assignment 1"
author: "Oguzhan Yetkin, Oleg Litvinov, Victor Retamal, group 4"
date: "15 March 2022"
output:
  pdf_document: default
  html_document:
    df_print: paged
highlight: tango
fontsize: 11pt
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

## Exercise 2. Airpollution.

The data were obtained to determine predictors related to air pollution. We want to investigate which explanatory variables need to be included into a linear regression model with oxidant as the response variable.

```{r}
pollution_df <- read.table("data/airpollution.txt", header=TRUE)
head(pollution_df)
```


### a)  Make some graphical summaries of the data. Investigate the problem of potential and influence points, and the problem of collinearity.

```{r}
# cor(pollution_df, method = c("spearman"))
if (!require("corrplot")) install.packages("corrplot")
library(corrplot)
par(mfrow=c(2, 1))
res <- cor(pollution_df, method='kendall')
corrplot(res, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45)
pairs(pollution_df, pch = 19)
```
```{r}
if (!require("psych")) install.packages("psych")
library(psych)
pairs.panels(pollution_df, 
             method = "spearman", # correlation method
             hist.col = "#00AFBB",
             density = TRUE,  # show density plots
             ellipses = TRUE # show correlation ellipses
             )
```


### b)  Use the added variable plot to depict the relationship between response oxidant and predictor wind. What is the meaning of the slope of fitted regression for this scatter plot?

```{r}
if (!require("car")) install.packages("car")
library(car)

attach(pollution_df)
mod = lm(oxidant~insolation+humidity+wind)
par(mfrow=c(2, 1))
avPlots(mod)
summary(mod)
```
The slopes on the plots reflect the regression coefficients from the original miltiple regression model mod. 
```{r}
attach(pollution_df)
y = residuals(lm(oxidant~insolation+humidity))
x = residuals(lm(wind~insolation+humidity))
plot(x, y, xlab='residual of wind', ylab='residual of oxidant')
```

### c)  Fit a linear regression model to the data. Use both the step-up and step-down methods to find the best model. If step-up and step-down yield two different models, choose one and motivate your choice. 

#### Step-up

```{r}

for(i in names(pollution_df)){
  if(i == 'oxidant'){next}
  # summary(lm(oxidant~i))
  print(summary(lm(paste('pollution_df$oxidant',  '~pollution_df$', i))))
}
```
Wind variable gives maximum increase in the R^2. The variable is significant. Therefore, we can continue.

```{r}
for(i in names(pollution_df)){
  if(i %in% c('oxidant', 'wind')){next}
  print(summary(lm(paste('pollution_df$oxidant',  '~pollution_df$wind+pollution_df$', i))))
}
```
Temperature variable works in the same way as the previous choice. Continue.

```{r}
for(i in names(pollution_df)){
  if(i %in% c('oxidant', 'wind', 'temperature')){next}
  print(summary(lm(paste('pollution_df$oxidant',  
                         '~pollution_df$wind',
                         '+pollution_df$temperature',
                         '+pollution_df$', 
                         i))))
}
```
Humidity has the highest R-squared increase but the variable is not significant. Therefore, we don't add it to the model. Resulting model is:
```{r}
print(summary(lm(pollution_df$oxidant~pollution_df$wind+pollution_df$temperature)))
```

oxidant = -5.2 - 0.4*wind + 0.5*temperature + error, with R-squared = 0.8.

#### Step-down

```{r}
summary(lm(pollution_df$oxidant~ pollution_df$wind + pollution_df$temperature + pollution_df$day + pollution_df$humidity + pollution_df$insolation))
```
Day has the largest p-value and the value is larger than 0.05. Removing it.

```{r}
summary(lm(pollution_df$oxidant ~ pollution_df$wind + pollution_df$temperature + pollution_df$humidity + pollution_df$insolation))
```
Insolation is the largers from the insignificant. Removing it.

```{r}
summary(lm(pollution_df$oxidant~ pollution_df$wind + pollution_df$temperature + pollution_df$humidity))
```
Humidity is the only insignificant. Removing.

```{r}
summary(lm(pollution_df$oxidant~ pollution_df$wind + pollution_df$temperature))
```
All remaining variables are significant. Resulting model: oxidant = -5.2 - 0.4*wind + 0.5*temperature + error, with R-squared = 0.8. The model is the same as obtained with the step-up approach.

d)  Determine 95% confidence and prediction intervals for oxidant using the model you preferred in c) for wind=33, temperature=54, humidity=77 and insolation=21.

```{r}
x1 <- pollution_df$wind
x2 <- pollution_df$temperature

mod = lm(pollution_df$oxidant ~ x1 + x2)

newxdata = data.frame(x1=33, x2=54)

predict(mod, newxdata, interval='prediction', level=0.95)
predict(mod, newxdata, interval='confidence', level=0.95)
```

